---
title: "get zip to district tables"
author: "Janet Young"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tigris)
library(tidyverse)
library(kableExtra)
library(sf) # for functions like st_join
library(tmap) # for plotting maps
## so we don't have to download big data multiple times
options(tigris_use_cache = TRUE)
```

I was having issues with memory usage, so I prepped R objects using code in (zip_to_districts_prepRobjects.R)(zip_to_districts_prepRobjects.R), using the `tigris` package to download data and the `sf` package to help subset regions according to overlaps.

Code looks like this (but I'm not running it here):
```{r download boundaries of various WA state features, eval=FALSE}
WA_state <- states(year=2020) %>% 
    filter(NAME=="Washington")
WA_counties <- counties(year=2020, state="WA")
King_county <- WA_counties %>% filter(NAME=="King")
WA_state_stateLegDists <- state_legislative_districts(year=2020, state="WA")
WA_state_congDists <- congressional_districts(year=2020, state="WA")
```

I used the `st_join` function from the `sf` package to figure out which regions overlap each other.  Result is a merged table (a bit like result of `left_join`). I'm not running it here, but code looks like this:
```{r show code for st_join, eval=FALSE}
zipCodes_congDists_joined <- st_join(WA_state_zipCodes, 
                                     WA_state_congDists, 
                                     join = st_intersects, 
                                     left=FALSE)
```

I turn those into plain two-column tibbles (zip-district) like this (again, not running it here):
```{r show code to make plain table, eval=FALSE}
zipCodes_congDists_joined_plain <- zipCodes_congDists_joined %>%
    select(ZCTA5CE20,NAMELSAD) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(NAMELSAD) %>% 
    rename(congressional_district=NAMELSAD) %>% 
    rename(zip_code=ZCTA5CE20)
```


The `join` argument to the `st_join` function controls how overlaps are calculated. I'm trying three different options (described [here](https://crd230.github.io/lab3.html#within)):
    + `join=st_intersects` if a zip code touches a district, we consider them together. Many zip codes touch >1 district
    + `join=st_within, largest=TRUE` - if a zip code touches >1 district, we choose the one with the largest overlapping geographic region
    + `join=st_within` a zip code must be totally within a district to be counted  
    
Let's use zip codes and congressional districts as an example to see how they behave (there's 10 congressional districts in WA, fewer than the 49 LDs).

Here I load in the joined tables I got in a separate R session:
```{r load congressional dist zip cod overlaps}
load("zip_to_districts_Robjects/zipCodes_by_CongDist_intersectLargest_plain.rda")
load("zip_to_districts_Robjects/zipCodes_by_congDist_intersect_plain.rda")
load("zip_to_districts_Robjects/zipCodes_by_congDist_within_plain.rda")
load("zip_to_districts_Robjects/zipCodes_by_stateLD_intersectLargest_plain.rda")
```

And now look at the results.

First look at results using the `join = st_intersects` option, show the first few lines:

```{r st_intersects understand how zip codes and congressional districts intersect}
zipCodes_by_congDist_intersect_plain %>% 
    head(6) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

This looks great.  We could export the `zipCodes_by_CongDist_intersect_plain` as a csv files and actually use it, but there's an issue we should deal with first: some zip codes are in >1 LD using the intersects method:  
```{r count num districts per zip code}
zipCodes_by_congDist_intersect_plain %>% 
    count(zip_code) %>% 
    rename(num_districts_per_zip=n) %>% 
    count(num_districts_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

Now look at results from the `join = st_within` option - now all districts are in 1 zip code. BUT there are a lot of zip codes missing - I think it's ignoring the ones in >1 district.

```{r count districts per zip code using st_within}
zipCodes_by_congDist_within_plain %>% 
    count(zip_code) %>% 
    rename(num_districts_per_zip=n) %>% 
    count(num_districts_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```


Which zip codes were in 3 districts when we use intersect? I'll show the first 8
```{r check out zip codes in 3 districts}
zipCodes_by_congDist_intersect_plain %>% 
    count(zip_code) %>% 
    filter(n==3) %>% 
    head(8) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```


One of those zip codes is 98040, which google maps tells me is Mercer Island

Mercer Island is only in dist 9 according to [wikipedia](https://en.wikipedia.org/wiki/Washington%27s_9th_congressional_district) and doesn't even touch 7 and 1

Might be something to do with how the lake is accounted for.   I suspect there's a way to deal with it - see [this doc](https://crd230.github.io/lab3.html#within)
```{r mercer island districts}
zipCodes_by_congDist_intersect_plain %>% 
    filter(zip_code=="98040")
```
Now look at results using `join = st_intersects, largest=TRUE` - I think that chooses the single region that has the biggest overlap.  

```{r try largest TRUE in st_join, eval=FALSE}
zipCodes_by_CongDist_intersectLargest_plain %>% 
    count(zip_code) %>% 
    rename(num_districts_per_zip=n) %>% 
    count(num_districts_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

Do the same for state LDs
```{r how many state LDs per zip code}

zipCodes_by_stateLD_intersectLargest_plain %>% 
    count(zip_code) %>% 
    rename(num_districts_per_zip=n) %>% 
    count(num_districts_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```


What happened to 98040 for Mercer Island? It's only in district 1. That's WRONG!  Should be 9. Does it contan getting a lot of the lake?

is it to do with the precision option I chose? try an even smaller number?

```{r mercer island now, eval=FALSE}
zipCodes_by_CongDist_intersectLargest_plain %>%
    filter(zip_code=="98040")
```


plot 98040
```{r plot 98040}
load("zip_to_districts_Robjects/WA_state_zipCodes.rda")
load("zip_to_districts_Robjects/WA_state_congDists.rda")
WA_state_zipCodes %>%
    filter(ZCTA5CE20=="98040") %>% 
    ggplot() +
    geom_sf()
```


```{r another plot}
tm_shape(WA_state_congDists) +
    tm_borders(col = "blue") +
    tm_text("CD116FP") +
    tm_shape( (WA_state_zipCodes %>% filter(ZCTA5CE20=="98040")) ) +  
    tm_borders(col = "red")
```


plot district 1 and 9


Sanity check - compare with Kelsey's list [here](https://docs.google.com/spreadsheets/d/17fAQh_PVPIS7fLTZNYDGqnYsQpe6Rdzl4SvsK5T-u0o/edit#gid=0)
```{r}

Kelsey_LD1 <- c("98011", "98012", "98021", "98034", "98036", "98043", "98072", "98077", "98296")

temp <- zipCodes_by_stateLD_intersectLargest_plain %>% 
    filter(congressional_district=="State Senate District 1") %>% 
    arrange(zip_code) %>% 
    mutate(inKelseyList = zip_code %in% Kelsey_LD1) 
temp %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

```{r zips in kelsey list not mine}
missingZips <- setdiff(Kelsey_LD1, temp$zip_code)
# [1] "98012" "98021" "98036"
zipCodes_by_stateLD_intersectLargest_plain %>% filter(zip_code %in% missingZips)

```

plot 98012 and the three districts it might overlap with. 21 is NOT the largest. this is WRONG

```{r another plot}
load("zip_to_districts_Robjects/WA_state_stateLegDists.rda")

tm_shape((WA_state_stateLegDists %>% 
              filter(SLDUST %in% c("021","001","044") ))) +
    tm_borders(col = "blue") +
    tm_text("SLDUST", col = "blue") +
    
    tm_shape( (WA_state_zipCodes %>% filter(ZCTA5CE20=="98012")) ) +  
    tm_borders(col = "red") +
    tm_text("ZCTA5CE20", col = "red") 
    
```

