---
title: "get zip to district tables"
author: "Janet Young"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tigris)
library(tidyverse)
library(kableExtra)
library(sf) # for functions like st_join
## so we don't have to download big data multiple times
options(tigris_use_cache = TRUE)
```

Use the `tigris` package to get the boundaries of various regions. This will take a few minutes.  

```{r download WA state boundaries}

WA_state_LDs <- state_legislative_districts("WA")
WA_state_CongDists <- congressional_districts("WA")
WA_state_ZipCodes <-  zctas("WA") 

#WA_counties <- counties("WA")
#WA_state_VotingDists <- voting_districts("WA")	

## something about this code seems to use a lot of memory, so I'm doing GC() quite often
gc(verbose = FALSE)

#### some stats on each object
# format(object.size(WA_counties), units = "auto")
#  2Mb
# format(object.size(WA_state_LDs), units = "auto")
# 2.8Mb
# format(object.size(WA_state_CongDists), units = "auto")
# 1.3Mb
# format(object.size(WA_state_VotingDists), units = "auto")
# 29 Mb
# format(object.size(WA_state_ZipCodes), units = "auto")
# 813.2Mb
# 
# dim(WA_counties)
# # [1] 39 18
# dim(WA_state_LDs)
# # [1] 49 13
# dim(WA_state_CongDists)
# # [1] 10 13
# dim(WA_state_VotingDists)	
# # [1] 7434   15
# dim(WA_state_ZipCodes) 
# # [1] 33791    10 
```

We can use the `st_join` function from the `sf` package to figure out which regions overlap each other.  Result is a merged table (a bit like result of `left_join`).

We can use `st_join` in two ways, using the `join` option (described [here](https://crd230.github.io/lab3.html#within)):
    + `join=st_intersects` if a zip code touches a district, we consider them together. Many zip codes touch >1 district
    + `join=st_within` a zip code must be totally within a district to be counted

Let's use zip codes and congressional districts as an example (there's 10 congressional districts in WA, compared to 49 LDs).

First let's try the `join = st_intersects` option and show the first few lines:

```{r st_intersects understand how zip codes and congressional districts intersect}
zipCodes_by_CongDist_intersect <- st_join(WA_state_ZipCodes, 
                                WA_state_CongDists, 
                                join = st_intersects, 
                                left=FALSE)
gc(verbose = FALSE)
# turn that into a tibble
zipCodes_by_CongDist_intersect_plain <- zipCodes_by_CongDist_intersect %>%
    select(ZCTA5CE20,NAMELSAD) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(NAMELSAD) %>% 
    rename(congressional_district=NAMELSAD) %>% 
    rename(zip_code=ZCTA5CE20)

zipCodes_by_CongDist_intersect_plain %>% 
    head(6) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

This looks great.  We could export the `zipCodes_by_CongDist_intersect_plain` as a csv files and actually use it, but there's an issue we should deal with first: some zip codes are in >1 LD using the intersects method:  
```{r count num districts per zip code}
zipCodes_by_CongDist_intersect_plain %>% 
    count(zip_code) %>% 
    rename(num_districts_per_zip=n) %>% 
    count(num_districts_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

Now let's try the `join = st_within` option - now all districts are in 1 zip code. BUT there are a lot of zip codes missing - I think it's ignoring the ones in >1 district.

```{r st_within understand how zip codes and congressional districts intersect}
zipCodes_by_CongDist_within <- st_join(WA_state_ZipCodes, 
                                WA_state_CongDists, 
                                join = st_within, 
                                left=FALSE)
gc(verbose = FALSE)
# turn that into a tibble
zipCodes_by_CongDist_within_plain <- zipCodes_by_CongDist_within %>%
    select(ZCTA5CE20,NAMELSAD) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(NAMELSAD) %>% 
    rename(congressional_district=NAMELSAD) %>% 
    rename(zip_code=ZCTA5CE20)
```
```{r count districts per zip code using st_within}
zipCodes_by_CongDist_within_plain %>% 
    count(zip_code) %>% 
    rename(num_districts_per_zip=n) %>% 
    count(num_districts_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```


Which zip codes were in 3 districts when we use intersect? I'll show the first 8
```{r check out zip codes in 3 districts}
zipCodes_by_CongDist_intersect_plain %>% 
    count(zip_code) %>% 
    filter(n==3) %>% 
    head(8) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```
One of those zip codes is 98040, which google maps tells me is Mercer Island

Mercer Island is only in dist 9 according to [wikipedia](https://en.wikipedia.org/wiki/Washington%27s_9th_congressional_district) and doesn't even touch 7 and 1

Might be something to do with how the lake is accounted for.   I suspect there's a way to deal with it - see [this doc](https://crd230.github.io/lab3.html#within)
```{r mercer island districts}
zipCodes_by_CongDist_intersect_plain %>% 
    filter(zip_code=="98040")
```

Let's try `largest=TRUE` in the `st_join` function - I think that chooses the single region that has the biggest overlap.  What happens?

It runs very slowly, and consumes a lot of memory, and is not finishing.  For now I won't run this code.   I want to test it either in a new R session, or perhaps with a smaller test dataset.

I can also probably help this along by using a separate R session to create and save the intersection objects

```{r try largest TRUE in st_join, eval=FALSE}
zipCodes_by_CongDist_intersectLargest <- st_join(WA_state_ZipCodes, 
                                WA_state_CongDists, 
                                join = st_intersects, 
                                largest=TRUE,
                                left=FALSE)
gc(verbose = FALSE)

# turn that into a tibble
zipCodes_by_CongDist_intersectLargest_plain <- zipCodes_by_CongDist_intersectLargest %>%
    select(ZCTA5CE20,NAMELSAD) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(NAMELSAD) %>% 
    rename(congressional_district=NAMELSAD) %>% 
    rename(zip_code=ZCTA5CE20)

zipCodes_by_CongDist_intersectLargest_plain %>% 
    count(zip_code) %>% 
    rename(num_districts_per_zip=n) %>% 
    count(num_districts_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```


What happened to Mercer Island?
```{r mercer island now, eval=FALSE}
zipCodes_by_CongDist_intersectLargest_plain %>%
    filter(zip_code=="98040")
```