---
title: "prep geographic overlaps"
author: "Janet Young"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {rmarkdown::render(inputFile,  encoding=encoding, output_dir="../reports") })
output: github_document
always_allow_html: true
---

Goal:  figure out the overlaps between various types of geographic regions.  Save a table, and make a plot for each LD.

We use the region data we saved earlier, using the [prep_geographic_areas.Rmd](Rscripts/prep_geographic_areas.Rmd) code.


```{r setup, include=FALSE}
library(tigris) # get map data
library(sf) # work with map data
library(tidyverse)
library(tmap) # for plotting maps
library(kableExtra) # for displaying tables
library(here) # helps manage file paths when working in an R project

options(tigris_use_cache = TRUE)

## we'll project into this coordinate system ("NAD83(2011) / Washington North", ID=6596)
chosenCoordReferenceSystem <- 6596

load(here("saved_R_objects/WA_state_zipCodes_proj.rda"))
load(here("saved_R_objects/WA_state_proj_shrink1km.rda"))
load(here("saved_R_objects/WA_stateLegDists_proj_shrink1km.rda"))
load(here("saved_R_objects/WA_congDists_proj_shrink1km.rda"))
load(here("saved_R_objects/exampleRegions_proj.rda" ))

## we will use the unshrunk versions when we plot
load(here("saved_R_objects/WA_stateLegDists_proj.rda"))
load(here("saved_R_objects/WA_congDists_proj.rda"))

### don't think I'm using these
# load(here("saved_R_objects/WA_state_proj.rda"))
# load(here("saved_R_objects/WA_counties_proj.rda"))
# load(here("saved_R_objects/WA_stateLegDists_proj.rda"))
# load(here("saved_R_objects/WA_congDists_proj.rda"))
```


Figure out the overlaps between zip codes and LDs. We keep the `sf` style object with the geographic information, but we also make a plain tibble.
```{r get LD to zip code mapping}
zip_LDshrink1km_intersects_proj <- st_join(WA_state_zipCodes_proj, 
                                           WA_stateLegDists_proj_shrink1km, 
                                           join = st_intersects, 
                                           left=FALSE)
zip_LDshrink1km_intersects_proj_plain <- zip_LDshrink1km_intersects_proj %>%
    select(ZCTA5CE20,NAMELSAD) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(NAMELSAD) %>% 
    rename(leg_dist=NAMELSAD) %>% 
    rename(zip_code=ZCTA5CE20) %>% 
    mutate(leg_dist_short=gsub("State Senate District ","LD_",leg_dist))
```

Do the same for congressional districts - get the overlaps between zip codes and CDs.
```{r get CD to zip code mapping}
zip_CDshrink1km_intersects_proj <- st_join(WA_state_zipCodes_proj, 
                                           WA_congDists_proj_shrink1km, 
                                           join = st_intersects, 
                                           left=FALSE)

zip_CDshrink1km_intersects_proj_plain <- zip_CDshrink1km_intersects_proj %>%
    select(ZCTA5CE20,NAMELSAD) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(NAMELSAD) %>% 
    rename(cong_dist=NAMELSAD) %>% 
    rename(zip_code=ZCTA5CE20) %>% 
    mutate(cong_dist_short=gsub("Congressional District ","CD_",cong_dist))
```


Now see which zip codes are in WA state, using the 1km 'shrunken' boundaries.

```{r zip codes in WA using 1km shrunken WA state}
zip_WAstate_shrink1km_intersects_proj <- st_join(WA_state_zipCodes_proj, 
                                           WA_state_proj_shrink1km, 
                                           join = st_intersects, 
                                           left=FALSE)

zip_WAstate_shrink1km_intersects_proj_plain <- zip_WAstate_shrink1km_intersects_proj %>%
    select(ZCTA5CE20) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(ZCTA5CE20) %>% 
    rename(zip_code=ZCTA5CE20)
```
The table is sorted by zip code. I'll show the top and bottom entries below.


```{r}
zip_WAstate_shrink1km_intersects_proj_plain %>% head(3) %>% kable() %>% kable_styling()
```
```{r}
zip_WAstate_shrink1km_intersects_proj_plain %>% tail(3) %>% kable() %>% kable_styling()
```

Looks good (mostly) - zip codes starting [980-994 are all in WA](https://simple.wikipedia.org/wiki/List_of_ZIP_Code_prefixes) (995 and above is Alaska)

Turns out 83856 is a weird zip code - it's mostly in Idaho but it DOES have a bit sticking out ~5 miles into WA state (north of Spokane).

Now we have table of all WA state zip codes, we'll be able to see whether there were any zip codes that did not get assigned to a district. 


How many LDs is each zip code in? boundaries don't line up, so some are in >1.  

```{r count num LDs per zip code}
zip_LDshrink1km_intersects_proj_plain %>% 
    mutate(zip_code2=factor(zip_code, 
                           levels=zip_WAstate_shrink1km_intersects_proj_plain$zip_code)) %>% 
    count(zip_code2, .drop=FALSE) %>% 
    rename(num_LDs_per_zip=n) %>% 
    count(num_LDs_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

What are the zips that were not assigned to LDs? They might be so small they got missed because of the 1km shrinkage of the LD boundaries
```{r show zips without LDs}
zips_without_LDs <- setdiff(zip_WAstate_shrink1km_intersects_proj_plain$zip_code,  
                            zip_LDshrink1km_intersects_proj_plain$zip_code)
zips_without_LDs
## I checked - there are no zips from outside WA state assigned to LDs 
# setdiff( zip_LDshrink1km_intersects_proj_plain$zip_code, zip_WAstate_shrink1km_intersects_proj_plain$zip_code)
```


How many CDs is each zip code in? boundaries don't line up, so some are in >1.  

```{r count num CDs per zip code}
zip_CDshrink1km_intersects_proj_plain %>% 
    mutate(zip_code2=factor(zip_code, 
                           levels=zip_WAstate_shrink1km_intersects_proj_plain$zip_code)) %>% 
    count(zip_code2, .drop=FALSE) %>% 
    rename(num_CDs_per_zip=n) %>% 
    count(num_CDs_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```
What are the zips that were not assigned to CDs? 
```{r show zips without CDs}
zips_without_CDs <- setdiff(zip_WAstate_shrink1km_intersects_proj_plain$zip_code,  
                            zip_CDshrink1km_intersects_proj_plain$zip_code)
zips_without_CDs
## I checked - there are no zips from outside WA state assigned to CDs 
## setdiff( zip_CDshrink1km_intersects_proj_plain$zip_code, zip_WAstate_shrink1km_intersects_proj_plain$zip_code)
```

Save all the R objects to use in other sessions
```{r save R objects}

save(zip_LDshrink1km_intersects_proj, file=here("saved_R_objects/zip_LDshrink1km_intersects_proj.rda"))
save(zip_LDshrink1km_intersects_proj_plain, file=here("saved_R_objects/zip_LDshrink1km_intersects_proj_plain.rda"))

save(zip_CDshrink1km_intersects_proj, file=here("saved_R_objects/zip_CDshrink1km_intersects_proj.rda"))
save(zip_CDshrink1km_intersects_proj_plain, file=here("saved_R_objects/zip_CDshrink1km_intersects_proj_plain.rda"))

save(zip_WAstate_shrink1km_intersects_proj, file=here("saved_R_objects/zip_WAstate_shrink1km_intersects_proj.rda"))
save(zip_WAstate_shrink1km_intersects_proj_plain, file=here("saved_R_objects/zip_WAstate_shrink1km_intersects_proj_plain.rda"))

save(zips_without_CDs, zips_without_LDs, file=here("saved_R_objects/missingZips.rda"))
```



Look at the range of number of zip codes per LD / CD
```{r count num districts per zip code}
numZipsPerLD <- zip_LDshrink1km_intersects_proj_plain %>% 
    count(leg_dist_short) %>% 
    rename(num_zips_per_LD=n) %>% 
    summarize(min_zips_per_LD=min(num_zips_per_LD),
              max_zips_per_LD=max(num_zips_per_LD)) %>% 
    t()

numZipsPerCD <- zip_CDshrink1km_intersects_proj_plain %>% 
    count(cong_dist_short) %>% 
    rename(num_zips_per_CD=n) %>% 
    summarize(min_zips_per_CD=min(num_zips_per_CD),
              max_zips_per_CD=max(num_zips_per_CD)) %>% 
    t() 

rbind(numZipsPerLD, numZipsPerCD) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

Now export the LD to zip tibble in a similar format to Kelsey's google sheet.  Can't just `pivot_wider()`, because some zips are present more than once and the columns have different lengths.  (or maybe we could if we messed around a bit)   I just used data.frame and a for loop instead.




```{r}

makeWideTable <- function(zipToDist_tbl, maxZips=0, 
                          missingZips=NULL, saveToFile=NULL) {
    x <- data.frame(row.names=1:maxZips)
    if(!is.null(missingZips)) {
        x[,"no_district_assigned"] <- NA
        x[1:length(missingZips),"no_district_assigned"] <- missingZips
    }
    distNames <- zipToDist_tbl %>% 
        select(dist) %>% 
        mutate(numeric_dist=as.integer(gsub("LD_|CD_","",dist))) %>% 
        arrange(numeric_dist) %>% 
        select(-numeric_dist) %>% 
        unique() %>% 
        unlist(use.names = FALSE) 
    for (thisDist in distNames) {
        theseZips <- zipToDist_tbl %>% 
            filter(dist==thisDist) %>% 
            select(zip_code) %>% 
            arrange(zip_code) %>% 
            unlist(use.names = FALSE)
        
        x[,thisDist] <- NA
        x[1:length(theseZips),thisDist] <- theseZips
    }
    if(!is.null(saveToFile)) {
        write.table(x, file=saveToFile, na="", quote=FALSE, sep="\t", row.names = FALSE)
    }
    return(x)  
}

### get LD zip table
LDzip_wide <- zip_LDshrink1km_intersects_proj_plain %>% 
    rename(dist=leg_dist_short) %>% 
    makeWideTable(maxZips=numZipsPerLD["max_zips_per_LD",1], 
                  missingZips=zips_without_LDs,
                  saveToFile=here("output_to_share/LD_zip_codes_uncurated_v1_2022_Oct30.tsv"))

### get CD zip table
CDzip_wide <- zip_CDshrink1km_intersects_proj_plain %>% 
    rename(dist=cong_dist_short) %>% 
    makeWideTable(maxZips=numZipsPerCD["max_zips_per_CD",1], 
                  missingZips=zips_without_CDs,
                  saveToFile=here("output_to_share/CD_zip_codes_uncurated_v1_2022_Oct30.tsv"))
```


now make a plot for each LD, showing the zip codes we assigned to it.

(will be able to modify this function to work on CDs too - need to fix some column names first)
```{r}

plotDistrictAndZipCodes <- function(allZipMaps=WA_state_zipCodes_proj,
                                    zipToDistrictPlainTbl,
                                    district_short_name,
                                    savePlot=NULL) {
    
    zips_to_plot <- zipToDistrictPlainTbl %>% 
        filter(leg_dist_short==district_short_name) %>% 
        select(zip_code) %>% 
        unlist(use.names = FALSE)
    myTitle <- paste(district_short_name, "(red) and assigned zip codes (uncurated)")
    
    p <- tm_shape(allZipMaps %>% filter(ZCTA5CE20 %in% zips_to_plot)) +
        tm_borders(col = "gray") +
        tm_fill(col="ZCTA5CE20", alpha=0.5,legend.show=FALSE) +
        tm_text("ZCTA5CE20", size=0.5) +
        
        tm_shape( WA_stateLegDists_proj %>% filter(shortName==district_short_name) ) +  
        tm_fill(col="pink", alpha=0.25) +
        tm_borders(col = "red", lty=2, lwd=2)  +
        
        tm_layout(main.title=myTitle, 
                  main.title.size = 0.9)
    if(!is.null(savePlot)) {
        # we don't overwrite
        if(!file.exists(savePlot)) {
            tmap_save(p, filename=savePlot)
        }
    }
    return(p)
}

### plot for a single LD to test the function
# p <- plotDistrictAndZipCodes(allZipMaps=WA_state_zipCodes_proj,
#                         zip_LDshrink1km_intersects_proj_plain,
#                         district_short_name="LD_1",
#                         savePlot = here("output_to_share/LD_zip_codes_uncurated_v1_2022_Oct30_plots/LD_1_uncurated_v1_2022_Oct30_plot.pdf")
#                         )

#### plots for all LDs
plotOutDir <- "output_to_share/LD_zip_codes_uncurated_v1_2022_Oct30_plots"
allPlots <- list()
for(thisLD in unique(zip_LDshrink1km_intersects_proj_plain$leg_dist_short)) {
    #cat("making plot for LD",thisLD,"\n")
    pdfFile <- paste(plotOutDir, "/", thisLD, "_uncurated_v1_2022_Oct30_plot.pdf",sep="" )
    pdfFile <- here(pdfFile)
    
    allPlots[[thisLD]] <- plotDistrictAndZipCodes(allZipMaps=WA_state_zipCodes_proj,
                        zip_LDshrink1km_intersects_proj_plain,
                        district_short_name=thisLD,
                        savePlot=pdfFile
                        )
}

## show one plot
allPlots[["LD_1"]]
```

```{r show package versions etc}
sessionInfo()
```