---
title: "zip_to_LD_use_2022_redistricted_boundaries"
author: "Janet Young"
date: "`r Sys.Date()`"
knit: (function(inputFile, encoding) {rmarkdown::render(inputFile,  encoding=encoding, output_dir="../reports") })
output: github_document
always_allow_html: true
---



Goals:   

1. read geography info for REDISTRICTED legislative and congressional districts into R

2. project coordinates for each set of regions into a more suitable coordinate system for our region ("NAD83(2011) / Washington North", ID=6596).  Code seems to behave better with this system, especially when trying to use 'buffer' functions to shrink areas a little bit.

3. get zip code overlaps, save as R objects

4. make plots and csv file as before

```{r setup, include=FALSE}
library(tigris) # get map data
library(sf) # work with map data
library(tidyverse)
library(tmap) # for plotting maps
library(kableExtra) # for displaying tables
library(here) # to help with file paths when using an Rproject

### functions and constants we'll use in >1 script are loaded from here (inluding definition of chosenCoordReferenceSystem):
source(here("Rscripts/shared_functions.R"))

load(here("saved_R_objects/WA_state_proj.rda"))
load(here("saved_R_objects/WA_state_zipCodes_proj.rda"))
load(here("saved_R_objects/zip_WAstate_shrink1km_intersects_proj_plain.rda"))

options(tigris_use_cache = TRUE)

```

Tell R where to find the WA state redistricting shape files
```{r file paths}
redist_data_dirs <- here("downloads/WA_redistricting_commission_2022_Oct30/Final_District_Shapes_2022_NAD_83/Final_District_Shapes_2022")
redist_data_dirs <- list.files(redist_data_dirs, full.names = TRUE)
redist_data_shpFiles <- list.files(redist_data_dirs, pattern=".shp$", full.names = TRUE)

names(redist_data_shpFiles) <-  tolower(sapply(strsplit(redist_data_dirs, "/"), function(x) {
    x[length(x)]
}))

```

Read them in
```{r read shp files}
CDs_redist <- st_read(redist_data_shpFiles["congessional"])
LDs_redist <- st_read(redist_data_shpFiles["legislative"])

# add a nicer name for plots
# I checked - the ID, DISTRICT and DISTRICTN columns of CDs_redist (and of LDs_redist) are identical to each other
CDs_redist$districtName <- paste("CD_", CDs_redist$ID, sep="")
LDs_redist$districtName <- paste("LD_", LDs_redist$ID, sep="")
```

Project coordinates to our chosen reference system, better for WA state
```{r project coords}
CDs_redist_proj <- projectCoords(CDs_redist)
LDs_redist_proj <- projectCoords(LDs_redist)
```


Now we 'shrink' the boundaries of each LD (or congressional district), by 1km, using the `st_buffer` function. This will give us better LD-to-zip mappings

```{r shrink regions}
CDs_redist_proj_shrink1km <- st_buffer(CDs_redist_proj, dist= -1000 )
LDs_redist_proj_shrink1km <- st_buffer(LDs_redist_proj, dist= -1000 )
```

Figure out the overlaps between zip codes and LDs. We keep the `sf` style object with the geographic information, but we also make a plain tibble.
```{r get LD to zip code mapping}
zip_LDredist_shrink1km_intersects_proj <- st_join(WA_state_zipCodes_proj, 
                                           LDs_redist_proj_shrink1km, 
                                           join = st_intersects, 
                                           left=FALSE)
zip_LDredist_shrink1km_intersects_proj_plain <- zip_LDredist_shrink1km_intersects_proj %>%
    select(ZCTA5CE20,districtName) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(ZCTA5CE20) %>% 
    arrange(districtName) %>% 
    rename(zip_code=ZCTA5CE20) 
```

Do the same for congressional districts - get the overlaps between zip codes and CDs.
```{r get CD to zip code mapping}
zip_CDredist_shrink1km_intersects_proj <- st_join(WA_state_zipCodes_proj, 
                                           CDs_redist_proj_shrink1km, 
                                           join = st_intersects, 
                                           left=FALSE)
zip_CDredist_shrink1km_intersects_proj_plain <- zip_CDredist_shrink1km_intersects_proj %>%
    select(ZCTA5CE20,districtName) %>%
    as_tibble() %>%
    select(-geometry) %>% 
    arrange(ZCTA5CE20) %>% 
    arrange(districtName) %>% 
    rename(zip_code=ZCTA5CE20) 
```




How many redistricted LDs is each zip code in? boundaries don't line up, so some are in >1.  

```{r count num LDs per zip code}
zip_LDredist_shrink1km_intersects_proj_plain %>% 
    mutate(zip_code2=factor(zip_code, 
                           levels=zip_WAstate_shrink1km_intersects_proj_plain$zip_code)) %>% 
    count(zip_code2, .drop=FALSE) %>% 
    rename(num_LDs_per_zip=n) %>% 
    count(num_LDs_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

Wow - one zip code is now in four LDs. It's 99301. It includes Pasco - take a quick look at which LDs it oerlaps:

```{r get the zip code in 4 LDs}
zipCodeWith4LDs <- zip_LDredist_shrink1km_intersects_proj_plain %>% 
    count(zip_code, .drop=FALSE) %>% 
    rename(num_LDs_per_zip=n) %>% 
    filter(num_LDs_per_zip==4) %>% 
    select(zip_code) %>% 
    unlist(use.names = FALSE)
zip_LDredist_shrink1km_intersects_proj_plain %>% 
    filter(zip_code == zipCodeWith4LDs) %>% 
    kable() %>% 
    kable_styling()
```



What are the zips that were not assigned to LDs? They might be so small they got missed because of the 1km shrinkage of the LD boundaries
```{r show zips without LDs}
zips_without_LDs <- setdiff(zip_WAstate_shrink1km_intersects_proj_plain$zip_code,  
                            zip_LDredist_shrink1km_intersects_proj_plain$zip_code)
zips_without_LDs
## I checked - there are no zips from outside WA state assigned to LDs 
# setdiff( zip_LDshrink1km_intersects_proj_plain$zip_code, zip_WAstate_shrink1km_intersects_proj_plain$zip_code)
```


How many CDs is each zip code in? boundaries don't line up, so some are in >1.  

```{r count num CDs per zip code}
zip_CDredist_shrink1km_intersects_proj_plain %>% 
    mutate(zip_code2=factor(zip_code, 
                           levels=zip_WAstate_shrink1km_intersects_proj_plain$zip_code)) %>% 
    count(zip_code2, .drop=FALSE) %>% 
    rename(num_CDs_per_zip=n) %>% 
    count(num_CDs_per_zip) %>% 
    rename(num_zip_codes=n) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```

What are the zips that were not assigned to CDs? 
```{r show zips without CDs}
zips_without_CDs <- setdiff(zip_WAstate_shrink1km_intersects_proj_plain$zip_code,  
                            zip_CDredist_shrink1km_intersects_proj_plain$zip_code)
zips_without_CDs
## I checked - there are no zips from outside WA state assigned to CDs 
## setdiff( zip_CDshrink1km_intersects_proj_plain$zip_code, zip_WAstate_shrink1km_intersects_proj_plain$zip_code)
```



Look at the range of number of zip codes per LD / CD
```{r count num districts per zip code}
numZipsPerLD <- zip_LDredist_shrink1km_intersects_proj_plain %>% 
    count(districtName) %>% 
    rename(num_zips_per_LD=n) %>% 
    summarize(min_zips_per_LD=min(num_zips_per_LD),
              max_zips_per_LD=max(num_zips_per_LD)) %>% 
    t()

numZipsPerCD <- zip_CDredist_shrink1km_intersects_proj_plain %>% 
    count(districtName) %>% 
    rename(num_zips_per_CD=n) %>% 
    summarize(min_zips_per_CD=min(num_zips_per_CD),
              max_zips_per_CD=max(num_zips_per_CD)) %>% 
    t() 

rbind(numZipsPerLD, numZipsPerCD) %>% 
    kable() %>% 
    kable_styling(font_size = 12)
```



Now export the LD to zip tibble in a similar format to Kelsey's google sheet.  Can't just `pivot_wider()`, because some zips are present more than once and the columns have different lengths.  (or maybe we could if we messed around a bit)   I made a function called `makeWideTable()` (see `Rscripts/shared_functions.R`) that uses data.frame and a for loop instead.

```{r export zip to district csv files}
### get LD zip table
LDzip_wide <- zip_LDredist_shrink1km_intersects_proj_plain %>% 
    rename(dist=districtName) %>% 
    makeWideTable(maxZips=numZipsPerLD["max_zips_per_LD",1], 
                  missingZips=zips_without_LDs,
                  saveToFile=here("output_to_share/version3_2022_Nov_redistricting2022_prettier/LDredist2022_zip_codes_uncurated_v3_2022_Nov2.tsv"))

### get CD zip table
CDzip_wide <- zip_CDredist_shrink1km_intersects_proj_plain %>% 
    rename(dist=districtName) %>% 
    makeWideTable(maxZips=numZipsPerCD["max_zips_per_CD",1], 
                  missingZips=zips_without_CDs,
                  saveToFile=here("output_to_share/version3_2022_Nov_redistricting2022_prettier/CDredist2022_zip_codes_uncurated_v3_2022_Nov2.tsv"))
```


now make a plot for each LD, showing the zip codes we assigned to it.

(will be able to modify this function to work on CDs too - need to fix some column names first)
```{r}

plotDistrictAndZipCodes_redist <- function(
        allZipMaps=WA_state_zipCodes_proj,
        allDistrictMaps=LDs_redist_proj,
        oldDistrictMaps=WA_stateLegDists_proj,
        zipToDistrictTbl=NULL,
        district_short_name,
        savePlot=NULL) {
    
    zips_to_plot <- zipToDistrictTbl %>% 
        filter(districtName==district_short_name) %>% 
        select(zip_code) %>% 
        unlist(use.names = FALSE)
    myTitle <- paste(district_short_name, "(2022 in red, pre-2022 in blue)\nand assigned zip codes (uncurated)")
    
    
    zipMapsThisDistrict <- allZipMaps %>% filter(ZCTA5CE20 %in% zips_to_plot)
    thisDistrictMap <- allDistrictMaps %>% filter(districtName==district_short_name)
    thisDistrictMapOld <- oldDistrictMaps %>% filter(shortName==district_short_name)
     
    wholeRegionToPlot <- st_union(st_union(zipMapsThisDistrict),st_union(thisDistrictMap)  )
    wholeRegionToPlot <- st_union(wholeRegionToPlot,st_union(thisDistrictMapOld) ) 
    p <- tm_shape(zipMapsThisDistrict, bbox=st_bbox(wholeRegionToPlot)) +
        tm_borders(col = "gray") +
        tm_fill(col="ZCTA5CE20", alpha=0.5,legend.show=FALSE) +
        tm_text("ZCTA5CE20", size=0.5) +
        
        ## pre-redistricting LD is in blue, no shading
        tm_shape( thisDistrictMapOld ) +  
        tm_borders(col = "cornflowerblue", lty=2, lwd=2)  +

        ## 2022 redistricted LD is in red, with shading
        tm_shape( thisDistrictMap ) +  
        tm_fill(col="pink", alpha=0.25) +
        tm_borders(col = "red", lty=2, lwd=2)  +
        
        tm_layout(main.title=myTitle, 
                  main.title.size = 0.9)
    if(!is.null(savePlot)) {
        # we don't overwrite
        if(!file.exists(savePlot)) {
            tmap_options(show.messages=FALSE)
            tmap_save(p, filename=savePlot)
            tmap_options_reset()
        }
    }
    return(p)
}

### load the pre-redistricting maps.  I won't do this at the top of the script, jsut to make absolutely sure that all the above analysis uses the new districts
load(here("saved_R_objects/WA_stateLegDists_proj.rda"))
load(here("saved_R_objects/WA_congDists_proj.rda"))

tmap_options(max.categories = as.integer(numZipsPerLD["max_zips_per_LD",1]))
# 
# ## plot for a single LD to test the function
# p <- plotDistrictAndZipCodes_redist(allZipMaps=WA_state_zipCodes_proj,
#         allDistrictMaps=LDs_redist_proj,
#         oldDistrictMaps=WA_stateLegDists_proj,
#         zip_LDredist_shrink1km_intersects_proj_plain,
#         district_short_name="LD_36",
#         savePlot = here("./test.pdf")
# )
# p


#### plots for all LDs
plotOutDir <- "output_to_share/version3_2022_Nov_redistricting2022_prettier/LDredist_zip_codes_uncurated_v3_2022_Nov2_plots"
allPlots <- list()
for(thisLD in unique(zip_LDredist_shrink1km_intersects_proj_plain$districtName)) {
    #cat("making plot for LD",thisLD,"\n")
    pdfFile <- paste(plotOutDir, "/", thisLD, "_redist_uncurated_v3_2022_Nov2_plot.pdf",sep="" )
    pdfFile <- here(pdfFile)
    
    allPlots[[thisLD]] <- plotDistrictAndZipCodes_redist(allZipMaps=WA_state_zipCodes_proj,
                        allDistrictMaps=LDs_redist_proj,
                        oldDistrictMaps=WA_stateLegDists_proj,
                        zipToDistrictTbl=zip_LDredist_shrink1km_intersects_proj_plain,
                        district_short_name=thisLD,
                        savePlot=pdfFile
                        )
}

## show one plot
allPlots[["LD_36"]]
```


```{r save R objects}
# not really needed - it's very quick to read in the shp files
save(LDs_redist,file=here("saved_R_objects/LDs_redist.rda"))
save(CDs_redist,file=here("saved_R_objects/CDs_redist.rda"))
save(LDs_redist_proj,file=here("saved_R_objects/LDs_redist_proj.rda"))
save(CDs_redist_proj,file=here("saved_R_objects/CDs_redist_proj.rda"))

save(LDs_redist_proj_shrink1km,file=here("saved_R_objects/LDs_redist_proj_shrink1km.rda"))
save(CDs_redist_proj_shrink1km,file=here("saved_R_objects/CDs_redist_proj_shrink1km.rda"))

save(zip_CDredist_shrink1km_intersects_proj,file=here("saved_R_objects/zip_CDredist_shrink1km_intersects_proj.rda"))
save(zip_CDredist_shrink1km_intersects_proj_plain,file=here("saved_R_objects/zip_CDredist_shrink1km_intersects_proj_plain.rda"))

save(zip_LDredist_shrink1km_intersects_proj,file=here("saved_R_objects/zip_LDredist_shrink1km_intersects_proj.rda"))
save(zip_LDredist_shrink1km_intersects_proj_plain,file=here("saved_R_objects/zip_LDredist_shrink1km_intersects_proj_plain.rda"))
```


Show R version and package versions, in case of troubleshooting

```{r show package versions etc}
sessionInfo()
```
